Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-01-10 main.tex
:pencil: Update contributions

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}

@@

	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and functions as an \emph{individual} phase shifter \cite{Wu2020}.
	% , accounting for a diagonal phase shift model with unit magnitude.
	The idea is extended to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	This allows \emph{cooperative} reflection --- wave impinging on one element can propagate within the circuit and depart partially from all elements.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	Channel shaping metrics can be classified into \emph{singular value centric} and \emph{power centric}.
	For diagonal \gls{ris}, the former has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023} in \gls{pc}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023} in \gls{ic}.
	The latter has been studied in terms of channel power \cite{Wu2019} in \gls{pc} and leakage interference \cite{Santamaria2023a} in \gls{ic}.
	When it comes to \gls{bd} \gls{ris}, the only metric analyzed is the channel power \cite{Nerini2023,Santamaria2023} while the \gls{mimo} case is still unexplored.\footnote{Single-stream \gls{mimo} with given precoder and combiner was considered in \cite{Nerini2023}. In terms of channel shaping, it is equivalent to \gls{siso}.}
	Although insightful, those attempts are far from complete and the methods are limited to specific scenarios.
	% This paper aims to quantify the channel shaping capability of asymmetric \gls{bd} \gls{ris} through theoretical analysis and numerical evaluation.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	% quantify the channel shaping capability of asymmetric \gls{bd} \gls{ris} through theoretical analysis and numerical evaluation.
	The contributions are summarized below.

	First, we examine the capability of \gls{bd} \gls{ris} to redistribute the channel singular values in \gls{mimo} \gls{pc}.
	The Pareto frontiers are characterized by optimizing the \emph{weighted sum of singular values}, where the weights can be positive, zero, or negative.
	It generalizes all singular value metrics and provides a powerful design framework.
	We then discuss analytical singular value bounds under rank-deficient and no-direct scenarios.

@@

	Second, we tailor a \gls{rcg} algorithm for asymmetric\footnote{Although the constraint of symmetric scattering parameters is widely respected in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2023a,Zhou2023,Li2023d,Bartoli2023}, it can be relaxed when the \gls{bd} \gls{ris} involves asymmetric passive components \cite{Ahn2006}, as previously assumed in \cite{Li2023b,Li2023c}. Symmetry can be enforced by projection if necessary.} \gls{bd} \gls{ris} optimization.
	Specifically, block-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold and concisely evaluated as matrix exponential.
	The proposed method not only features lower complexity and faster convergence than universal manifold optimization \cite{Absil2009,Pan2022d}, but also applies to any \gls{bd} \gls{ris} problem where gradient exists.
	% It is used to solve the Pareto singular problem, as well as rate maximization in \gls{mimo} \gls{pc} and \gls{wsr} maximization in \gls{mimo} \gls{ic}.
	% It effectively solves the Pareto singular problem.

	% , as well as rate maximization in \gls{mimo} \gls{pc} and \gls{wsr} maximization in \gls{mimo} \gls{ic}.

	% It helps to solve the aforementioned Pareto singular value problem.
	% We exploit it to solve the Pareto singular value problem and

	% It is exploited for the Pareto singular value problem, as well as rate maximization in \gls{mimo} \gls{pc} and \gls{wsr} maximization in \gls{mimo} \gls{ic}.
	% It is used to solve the Pareto singular value problem,

	Third, we consider power centric design
	% Second, we consider power centric design
	% exploit \gls{bd} \gls{ris} to power centric
	% maximize channel power in \gls{mimo} \gls{pc} and minimize leakage interference in \gls{mimo} \gls{ic}.

\begin{section}{Assumption}
	We introduce \gls{bd} \gls{ris} in \gls{mimo} \gls{pc} and \gls{ic}.
	% point-to-point and interference \gls{mimo} channels.
	All proposals are based on assumption of \emph{asymmetric} passive \gls{bd} \gls{ris}, i.e., symmetry constraint $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{T}$ is relaxed.
	This is feasible when asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006} are available.
	This assumption was also made in Hongyu's papers \cite{Li2023b,Li2023c}.
	For quadratic problems, the proposed algorithms may be extended to symmetric \gls{bd} \gls{ris} by replacing singular value decomposition with Takagi factorization \cite{Horn2012}.
\end{section}

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}

@@

	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and functions as an \emph{individual} phase shifter \cite{Wu2020}.
	% , accounting for a diagonal phase shift model with unit magnitude.
	The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	This allows \emph{cooperative} reflection --- wave impinging on one element can propagate within the circuit and depart partially from all elements.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	Channel shaping metrics can be classified into \emph{singular value centric} and \emph{power centric}.
	For diagonal \gls{ris}, the former has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023} in \gls{pc}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023} in \gls{ic}.
	The latter has been studied in terms of channel power \cite{Wu2019} in \gls{pc} and leakage interference \cite{Santamaria2023a} in \gls{ic}.
	For \gls{bd} \gls{ris}, the only metric analyzed is the channel power \cite{Nerini2023,Santamaria2023} while the \gls{mimo} case is still unexplored.\footnote{Single-stream \gls{mimo} with given precoder and combiner was considered in \cite{Nerini2023}. In terms of channel shaping, it is equivalent to \gls{siso}.}
	Although insightful, those attempts are far from complete and the techniques are limited to specific scenarios.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

	First, we examine the capability of \gls{bd} \gls{ris} to redistribute the channel singular values in \gls{mimo} \gls{pc}.
	The Pareto frontiers are characterized by optimizing the \emph{weighted sum of singular values}, where the weights can be positive, zero, or negative.
	It generalizes all singular value metrics and provides a powerful design framework.
	We then discuss some analytical singular value bounds under rank-deficient and no-direct scenarios.

@@

	Second, we tailor a \gls{rcg} algorithm for asymmetric\footnote{Although the constraint of symmetric scattering parameters is widely respected in the literature \cite{Shen2020a,Nerini2023,Santamaria2023,Fang2023,Nerini2023a,Zhou2023,Li2023d,Bartoli2023}, it can be relaxed when the \gls{bd} \gls{ris} involves asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, as previously assumed in \cite{Li2023b,Li2023c}. Symmetry can be enforced by projection if necessary.} \gls{bd} \gls{ris} optimization.
	Specifically, block-wise update is performed along the geodesics\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} of the Stiefel manifold and concisely evaluated as matrix exponential.
	The proposed method not only features lower complexity and faster convergence than universal manifold optimization \cite{Absil2009,Pan2022d}, but also applies to any \gls{bd} \gls{ris} problem where gradient exists.
	It effectively solves the Pareto singular problem.

	Third, we propose a closed-form iterative algorithm for power centric channel shaping problems.
	The idea is to successively approximate the quadratic objective by a sequence of affines and solve the local problems by \gls{svd}.
	Case studies are conducted for channel power maximization in \gls{pc} and leakage interference minimization in \gls{ic}.

	Fourth, we perform extensive simulations to validate and quantify the signal processing advantage of \gls{bd} \gls{ris}.
	Besides channel shaping, we also consider typical joint active and passive beamforming problems, including rate maximization in \gls{pc} and \gls{wsr} maximization in \gls{ic}.
	Results suggest that \gls{bd} \gls{ris} achieves a larger array gain in \gls{pc} and a higher \gls{dof} in \gls{ic}.
\end{section}

% \begin{section}{Assumption}
% 	We introduce \gls{bd} \gls{ris} in \gls{mimo} \gls{pc} and \gls{ic}.
% 	% point-to-point and interference \gls{mimo} channels.
% 	All proposals are based on assumption of \emph{asymmetric} passive \gls{bd} \gls{ris}, i.e., symmetry constraint $\mathbf{\Theta}_g = \mathbf{\Theta}_g^\mathsf{T}$ is relaxed.
% 	This is feasible when asymmetric passive components (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006} are available.
% 	This assumption was also made in Hongyu's papers \cite{Li2023b,Li2023c}.
% 	For quadratic problems, the proposed algorithms may be extended to symmetric \gls{bd} \gls{ris} by replacing singular value decomposition with Takagi factorization \cite{Horn2012}.
% \end{section}
